Rectangulotopes

### 2024-04-04 draft/rectangulotopes.tex
improve Section 3

% hello

@@

% Loday and anti-Loday accents (?)
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
% yin and yang accents (?)
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
% weak and strong equivalence on S_n
\newcommand{
\weakeq}{\asymp}
%\makeatletter
%\DeclareRobustCommand{\strongeq}{%
%  \mathbin{\mathpalette\scaletoplus@{\mkern1mu \smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}} \mkern1mu}}%
%}
%\newcommand{\scaletoplus@}[2]{%
%  \vcenter{%
%    \hbox{\sbox\z@{$\m@th#1+$}\resizebox{\wd\z@}{!}{$\m@th#1#2$}}%
%  }%
%}
%\makeatother
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}% % J: I would like to stack \backsim and \sim instead

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
We denote by %$i|j$ and $\frac{\ j\ }i$
$i \, \backslash \, j$ the situation where the rectangle labeled $i$ touches the rectangle labeled $j$ either from the left or from below in $R$. %, respectively.
We also use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise.

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

Given a rectangulation $R$, the two trees $T(R)$ and $S(R)$, where the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top,
define two partial orders on $[n]$.
We define the \defn{weak poset} $P_w(R)=([n],\prec_w)$ of a rectangulation $R$ as the transitive closure of the union of the two partial orders defined by the intertwining binary trees $T(R)$ and $S(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, the minimum of which is the root of $S(R)$ and the maximum is the root of $T(R)$.
We define the set $\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

\begin{theorem}
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $
\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item The congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of each congruence class are the twisted Baxter permutations,
  \item the maximal elements are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

\subsection{Loday and anti-Loday arcs}

@@

% jean: From this, do we directly have a description of the face lattice of $\WRP(n)$?}

\subsection{Shard polytopes of Loday and anti-Loday arcs}

Therefore, the quotientope is realized by a Minkowski sum of the shard polytopes defined by two families of arcs: the \defn{Loday arcs} and the \defn{anti-Loday arcs}, that go respectively above and below all elements of an interval $I$ of $[n]$.

\begin{lemma}
  \label{lem:lodaysp}
  Given a Loday arc $A$ defined by an interval $I$, the Loday shard polytope $\SP(A)$ is a translate of the standard simplex $\conv \set{ \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
  Given an anti-Loday arc $A$ defined by an interval $I$, the anti-Loday shard polytope $\SP(A)$ is a translate of the negative standard simplex $\conv \set{ - \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:weakMinkowski}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all Loday and anti-Loday shard polytopes.
  In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

\vincent{Say also the minkowski sum of all faces of the standard simplex corresponding to intervals, and all their opposite.}
\vincent{Say that this was already known by Law and Reading.}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f(S) \eqdef \#\bigset{ I\text{ interval of } [n] }{ I\cap S\not=\emptyset \text{ and } I\cap ([n]\ssm S) \not= \emptyset }.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

The proof relies on the following simple lemma stating that the submodular function of a Minkowski sum of deformed permutahedra is the sum of the submodular function of the summands.

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
\end{lemma}

\begin{proof}
  From \cref{lem:weakMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the Loday and anti-Loday shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. For a Loday arc $A$ defined by an interval $I$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \loday{f}_I(S) \eqdef \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, for an anti-Loday arc $A$ defined by an interval $I$, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \antiloday{f}_I(S) \eqdef - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f(S) \eqdef \sum_{I\text{ interval of }[n]} \loday{f}_I(S) + \antiloday{f}_I(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals intersecting both $S$ and its complement.
\end{proof}

@@

\begin{theorem}
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item The congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of each congruence class are the 2-clumped permutations,
  \item the maximal elements are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

>>>>

@@

% Loday and anti-Loday accents (?)
\makeatletter
\newcommand{\oset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits^{
    \vbox to#1{\kern-2\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
%
\newcommand{\uset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits_{
    \vbox to#1{\kern-7\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
\makeatother
%
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
\newcommand{\upArc}[1]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}}}
\newcommand{\downArc}[1]{\smash{\raisebox{-.05cm}{$\oset[.2ex]{#1}{\smile}$}}}
% yin and yang accents (?)
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
% weak and strong equivalence on S_n
\newcommand{
\weakeq}{\asymp}
%\makeatletter
%\DeclareRobustCommand{\strongeq}{%
%  \mathbin{\mathpalette\scaletoplus@{\mkern1mu \smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}} \mkern1mu}}%
%}
%\newcommand{\scaletoplus@}[2]{%
%  \vcenter{%
%    \hbox{\sbox\z@{$\m@th#1+$}\resizebox{\wd\z@}{!}{$\m@th#1#2$}}%
%  }%
%}
%\makeatother
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}% % J: I would like to stack \backsim and \sim instead

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
We denote by $i \, \backslash \, j$ the situation where the rectangle labeled $i$ touches the rectangle labeled $j$ either from the left or from below in $R$.
We also use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise.

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote nonempty and consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

Given a rectangulation~$R$, the source and target trees~$S(R)$ and~$T(R)$ define two partial orders on $[n]$ (the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top).
The \defn{weak poset}~$P_w(R)=([n],\prec_w)$ of the rectangulation~$R$ is the transitive closure of the union of the two partial orders defined by the intertwining binary trees~$S(R)$ and~$T(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$.
We define the set~$\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

\begin{theorem}[\cite{LawReading}]
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of each congruence class are the twisted Baxter permutations,
  \item the maximal elements are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

\subsection{Up and down arcs}

@@

\subsection{Shard polytopes of up and down arcs}

Therefore, the quotientope is realized by a Minkowski sum of the shard polytopes defined by two families of arcs: the \defn{up arcs}~$\upArc{I}$ and the \defn{down arcs}~$\downArc{I}$, that go respectively above and below all elements of an interval $I$ of $[n]$.

\begin{lemma}
  \label{lem:lodaysp}
%  Given an up arc~$\alpha$ defined by an interval $I$, the up shard polytope $\SP(\alpha)$ is a translate of the standard simplex~$\conv \set{ \b{e}_i }{ i\in I }$.
  The up shard polytope $\SP(\upArc{I})$ is a translate of the simplex~$\conv \set{ \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
%  Given a down arc~$\alpha$ defined by an interval $I$, the down shard polytope $\SP(\alpha)$ is a translate of the negative standard simplex~$\conv \set{ - \b{e}_i }{ i\in I }$.
  The down shard polytope $\SP(\downArc{I})$ is a translate of the negative simplex~$\conv \set{ - \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:weakMinkowski}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all up and down shard polytopes.
  In particular, it is the Minkowski sum of two opposite associahedra of J.-L.~Loday~\cite{LawReading}.
\end{lemma}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f_{\weakeq}(S) \eqdef \#\bigset{ I\text{ interval of } [n] }{ I \not\subseteq S \text{ and } I \not\subseteq [n]\ssm S }.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

%The proof relies on the following simple lemma stating that the submodular function of a Minkowski sum of deformed permutahedra is the sum of the submodular function of the summands.
%
%\begin{lemma}
%  \label{lem:submodsum}
%  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
%\end{lemma}

\begin{proof}
  %From \cref{lem:weakMinkowski,lem:submodsum},
  The submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the up and down shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. The shard polytope~$\SP(\upArc{I})$ of an up arc~$\upArc{I}$ is given by the submodular function
  \[
  f_{\upArc{I}}(S) \eqdef \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, the shard polytope~$\SP(\downArc{I})$ of a down arc~$\downArc{I}$ is given by the submodular function
  \[
  f_{\downArc{I}}(S) \eqdef - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f_{\weakeq}(S) = \sum_{I\text{ interval of }[n]} f_{\upArc{I}}(S) + f_{\downArc{I}}(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals contained neither in~$S$ nor in its complement.
\end{proof}

@@

\begin{theorem}[\cite{Reading}]
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of each congruence class are the 2-clumped permutations,
  \item the maximal elements are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}
